HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations

### 2021-05-03 docs/main.tex
remove latex math statement B\ast15 that prevents bibtex from compiling, comment out some highlights

  Our general approach for HAMdetector is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes. This Bayesian approach is advantageous because it allows use of: (1) prior information (e.g.\ knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\citep{Gelman2010}.

@@

Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise. These priors convey the a priori expectation that most coefficients in a regression model are close to 0, i.e.\ that non-zero coefficients are sparse. This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes. As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, though sometimes confounding factors such as HLA linkage disequilibrium may lead to the appearance of sizeable associations.
\hl{DaHo: kann man so nicht schreiben, weil die kompensatorischen Mutationen zu *echten* Assoziationen führen: ``and codon covariation,
e.g. in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.'' Ich finde trotzdem den Punkt der kompensatorischen Mutationen wichtig; sollte vielleicht in der Diskussion aufgegriffen werden. Z.B. ist es interessant zu sehen, ob unser Verfahren bei Positionen außerhalb von Epitopen immer noch einen Vorteil bei der Erkennung von Assoziationen liefert, oder zumindest solche Assoziationen wie in \citet{Ruhl2011} noch erkennt.}

@@

 For three of the false positives proposed by Fisher's exact test (Table~\ref{tab:false-positives}), HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03). One possible explanation for this observation is HLA linkage disequilibrium \hl{DaHo: and accordingly correlation between HLA alleles in the dataset due to random sampling variation. (Satz etwas opak.)} Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)). A similar observation can be made for K43R and D33E, although the correlation between the respective alleles is much weaker.

@@

Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L. With the simplest logistic regression model (model~1), both A*30 and B*13 have medium evidence of being associated with replacement P49L. However, with phylogeny and sparsity-promoting prior (model~3) both regression coefficients shrink close to 0 -- the associations are not convincingly supported by the data. Using epitope prediction as additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as likely associated with P49L. The association between P49L and A*30 (predicted by Fisher's exact test) remains shrunk towards 0.  
\section{Discussion}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\section*{Acknowledgements}
% \FloatBarrier
\section*{Funding}

@@

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{references}
\end{document}

>>>>

  Our general approach for HAMdetector is to fit Bayesian regression models that captures relationships between host HLA alleles and replacements in viral proteomes.

  This Bayesian approach is advantageous because it allows use of: (1) prior information (e.g.\ knowledge of effect magnitudes), (2) relevant additional information (phylogeny, epitope information), (3) a problem-specific structure, (4) partial pooling~\citep{Gelman2010}.

@@

Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise. These priors convey the a priori expectation that most coefficients in a regression model are close to 0, i.e.\ that non-zero coefficients are sparse. This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes. As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, with the exception of other factors such as HLA linkage disequilibrium and codon covariation, e.g. in the form of compensatory mutations that lie outside and epitope~\citep{Ruhl2011}.
% \hl{DaHo: kann man so nicht schreiben, weil die kompensatorischen Mutationen zu *echten* Assoziationen führen: ``and codon covariation,
% e.g. in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.'' Ich finde trotzdem den Punkt der kompensatorischen Mutationen wichtig; sollte vielleicht in der Diskussion aufgegriffen werden. Z.B. ist es interessant zu sehen, ob unser Verfahren bei Positionen außerhalb von Epitopen immer noch einen Vorteil bei der Erkennung von Assoziationen liefert, oder zumindest solche Assoziationen wie in \citet{Ruhl2011} noch erkennt.}

@@

 For three of the false positives proposed by Fisher's exact test (Table~\ref{tab:false-positives}), HAMdetector identifies associations with the same replacement but a different allele (P49L - B*13 instead of P49L - A*30; K43R - A*02 instead of K43R - B*13; and D33E - B*13 instead of D33E - A*03). One possible explanation for this observation is HLA linkage disequilibrium and accordingly correlation between HLA alleles in the dataset due to random sampling variation.

 %\hl{DaHo: and accordingly correlation between HLA alleles in the dataset due to random sampling variation. (Satz etwas opak.)} Indeed, out of the 12 times P49L is observed in sequences annotated with A*30, B*13 is also present in 5 of those cases (Spearman's rank correlation coefficient \(\rho = 0.5\)). A similar observation can be made for K43R and D33E, although the correlation between the respective alleles is much weaker.

@@

Figure~\ref{fig:linkage} shows regression coefficients of the HLA alleles A*30 and B*13 for replacement P49L. With the simplest logistic regression model (model~1), both A*30 and B*13 have medium evidence of being associated with replacement P49L. However, with phylogeny and sparsity-promoting prior (model~3) both regression coefficients shrink close to 0 -- the associations are not convincingly supported by the data. Using epitope prediction as additional source of information (model~4) allows to disentangle the association of the correlated alleles with P49L and identify B*13 as likely associated with P49L. The association between P49L and A*30 (predicted by Fisher's exact test) remains shrunk towards 0.

\subsection{HAMs outside epitopes}
\label{sec:hams-outs-epit}

imperfect epitope / processing predictions

compensatory mutations

we expect HAMs outside epitopes - are there such HAMs?

Figure

\section{Discussion}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
% %
% %     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
% %     as it may ignore the figures.
% %
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section*{Acknowledgements}
% \FloatBarrier
\section*{Funding}

@@

\bibliographystyle{natbib}
% \bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
% \bibliographystyle{bioinformatics}
%
% \bibliographystyle{plain}
%
\bibliography{references}
\end{document}
